Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2019-12-03 paper.Rnw
jp comment 21
moved text about rasch model from sec 5 (principal strat) to sec 2.4
(measuring hint usage)

adjusted text in sec 5

small change to dichotomization discussion in sec 4.1 (mediation)

For the current purpose, we narrow down the type
of hint data we model to align with our research focus and to avoid
some modeling pitfalls.
Since our posttest measured Algebra I skills, we only
considered worked problems from the Algebra I curriculum.
Next, although \citet{goldin2012learner} demonstrated the importance
of hint level, our focus is on the overall average effect of
requesting any hints; therefore, the variable measuring a student's
hint request on a problem was set one if the student requested any
hints on that problem, and zero otherwise.
% (More mundanely, we do not have access to data
% describing the number of hints available for every problem, which
% would hinder interpretation of any other measure of hint usage.)

@@

The observational study and (especially) the mediation analysis we
describe below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{\bar{h}_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.
To choose $c$, we fit a two-component modified Rasch mixture model
\citep{rasch} to hint request data, described in an online appendix.
The model classified approximately \Sexpr{round(mixProp*100)}\% of students as
high-hint users, so we chose $c$ as the
\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$,
approximately \Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))}.
This definition agreed with the model's classification in
approximately \Sexpr{round(agree*100)}\% of cases.

@@

In previous sections $\bar{h}$ was dichotomized into $H$; such
dichotomization is not necessary here.
However, $\bar{h}$ has some disadvantages as a continuous measure of
hint usage.
For one, since it is essentially a sample mean over a subset of each student's
worked problems, much of its variance is driven by the total number of
problems students worked---in particular, its extreme values belong to
those students who barely used the tutor at all.
Further, it does not account for varying difficulty of the tutor's
problems.
As an alternative, we fit a
modified Rasch model \citep[e.g.][]{rasch},
\begin{equation}\label{eq:rasch}
Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
\end{equation}
where $h_{ip}$ is the event that student $i$ requests a hint on
problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
section-level parameter for $s[p]$, the section that problem $p$ was
drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
function.
Typically, the student parameter $\eta_i$ measures student ability;
here, instead, it measures a student's proclivity to request a hint.
The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
dataset many problems were worked by very few students, so problem
level parameters would be hard to estimate.
Instead, we included the section-level parameter $\delta_{s[p]}$,
which measures the extent to which students tend to request hints on
problems in section $s$.
This essentially assumes that all problems in each section are equally
likely to elicit a hint.

We incorporated the model for hint requests, (\ref{eq:rasch}), into
the larger principal stratification model, replacing $\eta$ for
$\bar{h}$.
Then, the principal stratification estimand is
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
A full treatment of this variant of principal stratification, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

>>>>

For the current purpose, we narrow down the type
of hint data we model to align with our research focus.
Since our posttest measured Algebra I skills, we only
considered worked problems from the Algebra I curriculum.
Next, although \citet{goldin2012learner} demonstrated the importance
of hint level, our focus is on the overall average effect of
requesting any hints; therefore, the variable measuring a student's
hint request on a problem was set to one if the student requested any
hints on that problem, and zero otherwise.
% (More mundanely, we do not have access to data
% describing the number of hints available for every problem, which
% would hinder interpretation of any other measure of hint usage.)

@@

\subsubsection{Dichotomizing Hint Usage}\label{sec:dichotomizingHintUsage}
The observational study and mediation analysis below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{\bar{h}_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.

To choose $c$, we first fit a modified Rasch mixture model
\citep{rasch} to hint request data.
Specifically, we modeled the probability of a hint request in each problem
$p$, worked by student $i$, as
\begin{equation*}\label{eq:rasch}
Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
\end{equation*}
where $h_{ip}$ is the event that student $i$ requests a hint on
problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
section-level parameter for $s[p]$, the section that problem $p$ was
drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
function.
Typically, the student parameter $\eta_i$ measures student ability;
here, instead, it measures a student's proclivity to request a hint.
The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
dataset many problems were worked by very few students, so problem
level parameters would be hard to estimate.
Instead, we included the section-level parameter $\delta_{s[p]}$,
which measures the extent to which students tend to request hints on
problems in section $s$.
This essentially assumes that all problems within each section are equally
likely to elicit a hint.

Then, in order to dichotomize hint request behavior, we modeled
student effects $\eta$ as
\begin{equation}\label{eq:mixture}
\eta\sim
p_0\mathcal{N}(\mu_0,\sigma_0)+(1-p_0)\mathcal{N}(\mu_1,\sigma_1)
\end{equation}
and constrained $\mu_0<\mu_1$.
This mixture model clusters students into two categories: high hint
users have $\eta$ drawn from a normal distribution with mean $\mu_1$
and standard deviation $\sigma_1$, and low hint users have $\eta$
drawn from a normal distribution with mean $\mu_0$ and standard
deviation $\sigma_0$.

Our main parameter of interest here is $p_0$, the proportion of
students who are low-hint users.
We estimated $p_0\approx$\Sexpr{prettyNum(round(1-mixProp,2))},
classifying \Sexpr{round(mixProp*100)}\% of students as
high-hint users, so we chose $c$ as the
\Sexpr{round((1-mixProp)*100)}th percentile of $\bar{h}$,
approximately \Sexpr{sprintf('%.1f',quantile(hintAvgStud[['propHintStud']],1-mixProp))}.
This definition agreed with the model's classification (based on
$\eta$ as opposed to $\bar{h}$ in
approximately \Sexpr{round(agree*100)}\% of cases.

The Rasch model for hint useage \eqref{eq:rasch}, but not the mixture
model \eqref{eq:mixture} will play an important role in the principal
stratification analysis of Section \ref{sec:principalStratification}.

@@

In previous sections $\bar{h}$ was dichotomized into $H$; such
dichotomization is not necessary here.
However, $\bar{h}$ has some disadvantages as a continuous measure of
hint usage.
For one, since it is essentially a sample mean over a subset of each student's
worked problems, much of its variance is driven by the total number of
problems students worked---in particular, its extreme values belong to
those students who barely used the tutor at all.
Further, it does not account for varying difficulty of the tutor's
problems.
In section \ref{sec:dichotomizingHintUsage}, we showed that
dichotomized $\bar{h}$ largely agrees with a dicotomized version of a
more sophisticated measure of hint usage.
However, when modeling hint usage continuously, the correspondence may
not hold \citet[see][for a more complete discussion]{aoas}.

For those reasons, we modeled hint usage at the problem level, with
equation \eqref{eq:rasch}, but with two important differences.
First, a conceptual difference: the $\eta$ in \eqref{eq:rasch} was
replaced by $\eta (1)$, indicating that it is measuring
\emph{potential} hint usage---the hint usage a student would exhibit
were he or she assigned to the treatment condition.
Whereas $\eta$ is only defined for students assigned to the treatment condition,
$\eta(1)$ is defined for all the students in the study.
Second, a modeling difference: since we are not dichotomizing $\eta(1)$
in this analysis, we did not model it as a normal mixture, as in
\eqref{eq:mixture}.

We incorporated the model for hint requests, (\ref{eq:rasch}), into
the larger principal stratification model, replacing $\eta(1)$ for
$\bar{h}(1)$.
Then, the principal stratification estimand is
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
A full treatment of this variant of principal stratification, including a
discussion of identification and estimation, may be found in
\citet{aoas}.
